---
title: "Peer Assessment 1, Reproducable Research"
output: html_document
---

**This document contains the completed work for peer assignment #1 for the Coursera Reproducable Research class.**

###Loading & Preprocessing Data

---

A. I will download the dataset required for this assignment to my local work directory.  Description of this data from course website: 

*This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.*


The variables included in this dataset are:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA ) 
2. **date**: The date on which the measurement was taken in YYYY-MM-DD format 
3. **interval**: Identifier for the 5-minute interval in which measurement was taken

B. I will read this data into a dataframe and inspect the structure of the data
```{r}
activityData<-read.csv('activity.csv', header = TRUE,as.is = FALSE)
```


###Histogram, Mean & Median

---

A. I will make a new data frame with sum of steps day by day:
```{r}
activityEachDay<-aggregate(steps~date,activityData,sum,na.rm=TRUE)
activityEachDay$date<-as.POSIXlt(activityEachDay$date,"%Y-%m-%d")
names(activityEachDay)[2]<-"sum_steps"
```

B.Make The Histogram:


```{r}
hist(activityEachDay$sum_steps,
     col = "red",
     main = "Histogram of the Total Number of Steps Each Day",
     xlab = "Total Number of Steps Each Day",
     breaks = 10)
```
C.Report the mean and median values:

```{r}
mean(activityEachDay$sum_steps)
median(activityEachDay$sum_steps)
```


###Daily Activity Pattern

----------------------------


**Next, I will try to visualize the daily activity pattern**

The below plot addresses the following items:

A. Calculate the mean number of steps taken across all the days
```{r}
activityEachInterval<-aggregate(steps~interval,activityData,mean,na.rm=TRUE)
names(activityEachInterval)[2]<-"mean_steps"
)

```
B.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
plot(
        x = activityEachInterval$interval,
        y = activityEachInterval$mean_steps,
        type = "l",
        main = "The Average Number of Steps for Each 5-Min Interval",
        xlab = "5-Minute Interval",
        ylab = "Average Number of Steps"
)

```
C.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
activityEachInterval[activityEachInterval$mean_steps==max(activityEachInterval$mean_steps,na.rm = TRUE),]
```
---

###Missing Values

A. Calculate & Report The Number of Missing Values
```{r}
nrow(activityData[is.na(activityData$steps),])
```

B. Devise a strategy for filling in all of the missing values in the dataset. I will use the mean for that 5-minute interval to replace missing values.I will create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activityNoNA<-merge(activityData,activityEachInterval,all.y=FALSE, all.x = TRUE, by = "interval",sort = FALSE)
activityNoNA[is.na(activityNoNA$steps),"steps"]<-as.integer(round(activityNoNA$mean_steps[is.na(activityNoNA$steps)]))
activityNoNA<-activityNoNA[,c("date","interval","steps")]
activityNoNA<-activityNoNA[order(activityNoNA$date,activityNoNA$interval),]
rownames(activityNoNA)<-c()
```

C.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}

#### Calculate the total number of steps taken each day with No NA data frame and store the data in a new dataframe

activityNNEachDay<-aggregate(steps~date,activityNoNA,sum)
activityNNEachDay$date<-as.POSIXlt(activityNNEachDay$date,"%Y-%m-%d")
names(activityNNEachDay)[2]<-"sum_steps"

##Make a histogram to show steps taken for each day

hist(activityNNEachDay$sum_steps,
     col = "red",
     main = "Histogram of the Total Number of Steps Each Day (NA romoved)",
     xlab = "Total Number of Steps Each Day",
     breaks = 10)

## Calculate mean and median number of steps taken each day with No NA data frame

mean(activityNNEachDay$sum_steps)
median(activityNNEachDay$sum_steps)

```

**Answer To Question:**
The mean and the median are now almost the same after replacing missing values with the mean value for the relevant interval. Because I'm useing the mean value for the replacements, it should not influence the results much.

###Are there differences in activity patterns between weekdays and weekends?

---
A.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
activityNoNA$date<-as.POSIXlt(activityNoNA$date,"%Y-%m-%d")
activityNoNA$daytype<-"weekday"
activityNoNA[weekdays(activityNoNA$date, abbreviate = TRUE) %in% c("Sat","Sun"),"daytype"]<-"weekend"
activityNoNA$daytype<-as.factor(activityNoNA$daytype)
```
B. Calculate the mean number of steps taken each 5-min interval for weekdays and weekends saperately and store the data in a new dataframe
```{r}
activityNNEachInterval<-aggregate(steps~interval + daytype, data = activityNoNA, mean)
names(activityNNEachInterval)[3] <- "mean_steps"
```
C.Plot the activity patterns for weekdays and weekends with lattice package
```{r}
library(lattice)
xyplot(
        mean_steps ~ interval | daytype,
        activityNNEachInterval,
        type = "l",
        layout = c(1,2),
        main = "The Average Number of Steps for Each 5-Min Interval during Weekday Days or Weekend Days",
        xlab = "5-Minute Interval",
        ylab = "Average Number of Steps"
)
```
**Answer To Question:**
The weekdays and weekends patterns are slight different.The major difference is that during the weekends, the max number of steps taken in 5-min interval is much smaller than weekdays. The peak for weekdays could be caused by walking to the work place.

